Categorical Foundations of E-Graphs With Binding

### 2025-07-07 POPL2026/notes/SlottedNotes.tex
notes and popl template

\input{../preamble}
\input{../macros}
\input{../sample.tikzstyles}
\input{../hypergraph.tikzstyles}
\input{../hypergraph.tikzdefs}

@@

while such rules are absorbed when using string diagrams (or hypergraphs) akin to proof nets, they are not absorbed by e-graph representation.
Below one could see the number of e-nodes for different set of equations enabled for saturating an expression \texttt{plus 2 2} using Church encoding.

\begin{tabular}{lcccc}
  Set of equations & built-in subst & explicit subst & explicit subst & explicit subst\\
  &&& + interchange&  + interchange\\
  &&&&  + comm\\
  \# of e-nodes & 40 & 161 & 289 & 356
\end{tabular}

@@

\bibliographystyle{acm}
\bibliography{../bibliography}

>>>>

\input{../../preamble}
\input{../../macros}
\input{../../sample.tikzstyles}
\input{../../hypergraph.tikzstyles}
\input{../../hypergraph.tikzdefs}

@@

while such rules are absorbed when using string diagrams (or hypergraphs) akin to proof nets, they are not absorbed by e-graph representation.
In Table~\ref{tbl:slotted} one could see the number of e-nodes for different set of equations enabled for saturating an expression \texttt{plus 2 2} using Church encoding.

\begin{table}
\begin{tabular}{lcccc}
  Set of equations & built-in subst & explicit subst & explicit subst & explicit subst\\
  &&& + interchange&  + interchange\\
  &&&&  + comm\\
  \# of e-nodes & 40 & 161 & 289 & 356
\end{tabular}
\caption{Slotted e-graphs benchmarks}
\label{tbl:slotted}
\end{table}

@@

Recall that by the virtue of free enrichment we have the equality $\Lambda(f + g) = \Lambda(f) + \Lambda(g)$.
The intuitive interpretation of this equality is that $f \sim g$ (due to rewrite rules being applied) implies $\Lambda(f) \sim \Lambda(g)$.
This is exactly one of the equivalences that a slotted e-graph from~\cite{slotted-egraphs} maintains.

\[
\text{bind}\;\$s_1\; x \cong \text{bind}\;\$s_2\;y \;\; \text{if} (\$s_1 \$f) * x \cong (\$s_2 \$f) * y, \text{where $f$ is fresh}
\]

\bibliographystyle{acm}
\bibliography{../../bibliography}

### 2025-07-07 POPL2026/paper/sections/categorical_semantics.tex
notes and popl template

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Figure \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.
Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.
% Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

>>>>

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Fig. \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.
Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.
% Fig. \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

### 2025-07-07 POPL2026/paper/sections/soundness.tex
notes and popl template

\begin{definition}
\label{def:rewrite}
    We say that a $\Sigma$-term $f$ rewrites to a $\Sigma$-term $g$ modulo SMC laws via a rewrite rule $\langle l, r \rangle$ (notation, $\leadsto_{\langle l, r \rangle}$) if they are representable as
    $
    f = c_{1};(id_{k} \otimes l);c_{2} $ and $ g = c_{1};(id_{k} \otimes r);c_{2}
    $.
\end{definition}
This notion generalises to rewriting of closed $\Sigma$-terms
\begin{definition}
    For closed-$\Sigma$-terms $f,g$, $f \leadsto_{\langle l, r \rangle}$ modulo SMC laws if they are representable as either in Definition~\ref{def:rewrite}, or as 
    $
    f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
    $
    and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
\label{lemma:normal_form}
    Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
    $
    f_{1} + \ldots + f_{n}
    $ modulo SMC laws and $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using distributivity law~\ref{law:distributivity}.
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
    For closed $\Sigma^{+}$-term $f,g$, $f \leadsto_{\langle l, r \rangle}$ SMC laws, $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} if they are representable as
    $
    f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
    $
    such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Next, to formulate the correspondence between rewriting systems we first define interpretation $\llbracket - \rrbracket$ of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of operators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in Appendix~\ref{sec:appendix:interpretation}.
The only missing cases, $\llbracket \textsf{ev}_{A,B} \rrbracket$ and $\llbracket \Lambda_{A,B,C} \rrbracket$, are given in Fig.~\ref{fig:ev_and_lambda}.

@@

Then we can make $\WellTypedMdaEcospans$ into $\catname{SLat}$-category by defining a join of two cospans and introducing DPOI rewrite rules to satisfy the laws of $\catname{SLat}$ symmetric monoidal category.
We also add a structural rewrite schema rule corresponding to distributivity law~\ref{law:distributivity}.
The join of two cospans is defined as in Fig.~\ref{fig:f+g} and the quotienting is done through DPOI rewrite schema rules, one for each axiom.
For example, the distributivity rule for $\otimes$ is given by the following schema rule
\[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/semilattice_rule_1}~.
}
\]

@@

This explicit quotienting allows us to only consider cospans of the form $f_1 + \ldots + f_{n}$, similar to Lemma~\ref{lemma:normal_form}, such that carrier of $f_{i}$ does not contain any hierarchical edges from $\textcolor{red}{E}$.

@@

\begin{proposition}[Proposition 26 and Proposition 27~\cite{fscd}]
\label{prop:fscd}
For two closed $\Sigma$-terms $f$ and $g$ and a $\mathcal{E}$ equation $l = r$  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ where the latter is defined in a subcategory of $\WellTypedMdaEcospans$ where carriers only contain edges from $\colorbox{yellow}{E} \cup \textcolor{closed-color}{E}$.
\end{proposition}

\begin{proposition}
    For two closed $\Sigma^{+}$-terms $f$ and $g$ and a $\mathcal{E}$ equation $l = r$  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
By definition, we have 
$
  f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
such that $f_{i} \leadsto f_{j}$ for some $i, j$.
This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{red}{E}$, we can apply Proposition~\ref{prop:fscd}.
\end{proof}

>>>>

\begin{definition}
\label{def:rewrite}
    We say that a $\Sigma$-term $f$ rewrites to a $\Sigma$-term $g$ modulo SMC laws via a rewrite rule $\langle l, r \rangle$ (notation, $\leadsto_{\langle l, r \rangle}$) if they are representable as
    $
    f = c_{1};(id_{k} \otimes l);c_{2} $ and $ g = c_{1};(id_{k} \otimes r);c_{2}
    $.
\end{definition}
This notion generalises to rewriting of closed $\Sigma$-terms
\begin{definition}
    For closed-$\Sigma$-terms $f,g$, $f \leadsto_{\langle l, r \rangle} g$ modulo SMC laws if they are representable as either in Definition~\ref{def:rewrite}, or as 
    $
    f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
    $
    and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
\label{lemma:normal_form}
    Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
    $
    f_{1} + \ldots + f_{n}
    $ modulo SMC laws and $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using distributivity law~\ref{law:distributivity}.
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
    For closed $\Sigma^{+}$-term $f,g$, $f \leadsto_{\langle l, r \rangle}$ SMC laws, $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} if they are representable as
    $
    f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
    $
    such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Next, to formulate the correspondence between rewriting systems we first define interpretation $\llbracket - \rrbracket$ of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of operators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in Appendix~\ref{sec:appendix:interpretation}.
The only missing cases, $\llbracket \textsf{ev}_{A,B} \rrbracket$ and $\llbracket \Lambda_{A,B,C} \rrbracket$, are given in Fig.~\ref{fig:ev_and_lambda}.

@@

Then we can make $\WellTypedMdaEcospans$ into $\catname{SLat}$-category by defining a join of two cospans and introducing DPOI rewrite rules to satisfy the laws of $\catname{SLat}$ symmetric monoidal category.
We also add a structural rewrite schema rule corresponding to distributivity law~\ref{law:distributivity}.
The join of two cospans is defined as in Fig.~\ref{fig:f+g} and the quotienting is done through DPOI rewrite schema rules, one for each axiom.
For example, the distributivity rule for $\otimes$ is given by the following schema rule
\ifdefined\ONECOLUMN
\[
\adjustbox{width=0.7\linewidth}{
\tikzfig{./figures/semilattice_rule_1}~.
}
\]
\else
\[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/semilattice_rule_1}~.
}
\]
\fi

@@

This explicit quotienting allows us to only consider cospans of the form $f_1 + \ldots + f_{n}$, similar to Lemma~\ref{lemma:normal_form}, such that carrier of $f_{i}$ does not contain any hierarchical edges from $\textcolor{e-color}{E}$.

@@

\begin{proposition}[Proposition 26 and Proposition 27~\cite{fscd}]
\label{prop:fscd}
For two closed $\Sigma$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$, where the latter is defined in a subcategory of $\WellTypedMdaEcospans$ where carriers only contain edges from $\colorbox{yellow}{E} \cup \textcolor{closed-color}{E}$.
\end{proposition}

\begin{proposition}
    For two closed $\Sigma^{+}$-terms $f$ and $g$ and an $\mathcal{E}$ equation $l = r$,  $f \leadsto_{\langle l, r \rangle} g$ if and only if $\llbracket f \rrbracket \Rrightarrow_{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.
\end{proposition}
\begin{proof}
By definition, we have 
$
  f = f_{1} + \ldots + f_{n} $ and $  g = f_{1} + \ldots + f_{n}$,
such that $f_{i} \leadsto f_{j}$ for some $i, j$.
This yields two cospans of the form $\llbracket f_{1} \rrbracket + \ldots + \llbracket f_{n} \rrbracket$ for $\llbracket f \rrbracket$ and $\llbracket g \rrbracket$ and, by noting that every $f_{i}$ is a $\Sigma$-term, and carriers in every $\llbracket f_{i} \rrbracket$ have no edges from $\textcolor{e-color}{E}$, we can apply Proposition~\ref{prop:fscd}.
\end{proof}
